A general and efficient representation of Ancestral Recombination Graphs

### 2023-08-31 paper.tex
Wilders comments

% Middle Authors
\author[4]{Gregor Gorjanc}
\author[5]{Anthony W. Wohns}

@@

\affil[1]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, OX3 7LF, UK}
\affil[2]{Department of Statistics, University of Oxford, OX1 3LB, UK}
\affil[3]{Department of Statistics, University of Warwick, CV4 7AL, UK}
\affil[4]{The Roslin Institute and Royal (Dick) School of Veterinary Studies, University of Edinburgh, EH25 9RG, UK}
\affil[5]{Broad Institute of MIT and Harvard, Cambridge, MA 02142, USA}

@@

% What do we do?
Here we precisely define an encoding for ARGs based on specific genomes
and their intervals of genetic inheritance. We contrast this ``gARG''
encoding with the classical approach derived from the coalescent
with recombination stochastic process of Griffiths
and colleagues~\citep{griffiths1991two,ethier1990two,
griffiths1996ancestral,griffiths1997ancestral}. As this is based
on recording evolutionary events, we refer to it as the event ARG, or
``eARG'' encoding.
We show that the gARG encoding is both flexible and efficient,
and can represent varying levels of precision about historical
events.
[MORE]

@@

\section*{Genome ARGs}
% There is no fundamental reason, however, that the data structure we use
% to describe patterns of genetic inheritance must be in terms
% of specific causal events. As demonstrated in subsequent sections,
% there are significant advantages to using
% a slightly different formulation, which
% we call a ``genome ARG'' (gARG).
In this section we define a concrete definition of an ARG in terms
of particular genomes and their patterns of inheritance, which
we refer to as a ``genome ARG``, or gARG.
First, let us define a genome as the complete set of genetic material that a child
inherits from one parent (i.e., haploids have one genome, diploids two, etc.).
Then, a gARG is a graph in which nodes represent
monoploid genomes  and edges represent
genetic inheritance between an ancestor and a descendant,
where each edge is annotated with the coordinates
of the genomic intervals over which genetic inheritance occurs.
We can therefore view a gARG for a given set of individuals
as a subset of their pedigree, but, crucially,
where each individual is represented by two nodes (for diploids).
% To recover the local tree for a genomic
% position $x$, we traverse the graph rootwards from the leaves
% (as in an eARG) and at a particular node, we follow the outbound
% edge associated with the genomic interval containing $x$.
% Recent discussions have illustrated the
% relationship between ARGs and
% pedigrees~\citep{mathieson2020ancestry,brandt2021evaluation},
% and the deep connections between the structures are
% well understood
% \citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
% speed2015naturereviewsgenetics}.
This is illustrated in Fig.~\ref{fig-arg-in-pedigree},
which
shows how the genomes embedded in an example (highly inbred)
pedigree correspond to the nodes in a gARG.
In this diploid example
a genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination.

@@

There are some interesting consequences to these definitions, exemplified
in Figure~\ref{fig-arg-in-pedigree}. Firstly,
there is
is no limitation on mating system, ploidy, or age structure. By making the
basic unit of analysis an individual's \emph{genomes} (i.e., two per individuals
for diploids), any pattern of inheritance among mixed ploidy can be
accommodated.
Secondly, because we only record the \emph{intervals} of inheritance
between genomes (and not the causal \emph{processes}),
arbitrary patterns of genetic inheritance such as
gene conversions,
multiple recombinations
across different chromosomes~\citep{fearnhead2003ancestral,koskela2019robust},
and so on can be directly expressed.

Thirdly, ``local`` trees that we generate for a particular position on the
genome are not explicitly concerned with ``coalescence'' (nodes at which
one or more samples find a common ancestor); for example, node $e$
is unary in the left-most tree of Figure~\ref{fig-arg-in-pedigree},
denoting the fact that $e$ the ancestor of $a$ over the interval $(0, 2]$,
but no coalescence occurs in this interval (see section XXX for more
discussion). Fourth, recombination is modelled in terms of
its outcomes in terms of inheritance intervals among parental genomes;
thus, we will have a child node (the recombinant; e.g. node $a$
Figure~\ref{fig-arg-in-pedigree}) and two (or more; see section XXX)
paternal genomes which this child has inherited segments from.
Finally, gARGs are not defined in terms of the
coalescent with recombination (or any other) stochastic process.
Thus, the gARG encoding is focused on faithfully representing the
\emph{outcome} of the evolutionary processes of genetic inheritance and
recombination, and is agnostic to the details of those processes.

@@

The gARG representation, on the other hand,
can be either retrospective or \emph{prospective}.
The edges in a gARG represent a path of genetic inheritance from
ancestor to descendant through some
number of generations (ultimately from cell to cell),
and are annotated
with the genome coordinates over which inheritance occurs.
In a prospective gARG, the inheritance intervals $I$
on an edge joining ancestor genome $v$ and descendant $u$
defines \emph{all} regions of genome that $u$ inherited from $v$.
Thus, these  inheritance intervals are
concerned only with the genetic relationship
between the nodes in question, and are not defined with respect
to any given set of samples. We are simply recording the
direct genetic inheritance between two genomes
(in the simplest case, between a parent and a child).

@@

The remaining nodes are MRCAs of some subset of the samples
at \emph{some} positions along the genome. We still have
some unary nodes in the local trees, but these nodes will
correspond to a coalescence in at least one other
local tree. For example, node  \noderef{k} is unary in the second tree
of Fig.~\ref{fig-simplification}C, but is either binary
or ternary in all subsequent trees (recall this is a Wright-Fisher
simulation). The final level of simplification is to alter the edge annotations
such that, although no nodes are removed from the graph, all
unary nodes disappear from the local trees (Fig.~\ref{fig-simplification}D).
Note that although this last stage produces simpler local trees, by
removing information about the exact paths taken by lineages through
the graph, we lose potentially useful information about shared edges
between trees. The extent to which this information can be inferred
and utilised is an interesting open question.

@@

\section*{Discussion}
Recent breakthroughs have made ARG inference methods practical
for the first time, and there has been a surge of interest
in inference methods, their evaluation and application.
The prospect of ARGs being used routinely in population
and statistical genetics applications is tantalising,
but in reality there is  substantial work to be done to
enable this, and much remains to be done.

>>>>

% Middle Authors
\author[4]{Gregor Gorjanc}
\author[5,6]{Anthony W. Wohns}

@@

\affil[1]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, OX3 7LF, UK}
\affil[2]{Department of Statistics, University of Oxford, OX1 3LB, UK}
\affil[3]{Department of Statistics, University of Warwick, CV4 7AL, UK}
\affil[4]{The Roslin Institute and Royal (Dick) School of Veterinary Studies, University of Edinburgh, EH25 9RG, UK}
\affil[5]{Broad Institute of MIT and Harvard, Cambridge, MA 02142, USA}
\affil[6]{Department of Genetics, Stanford University School of Medicine, Stanford, CA 94305, USA}

@@

% What do we do?
Here we precisely define an encoding for ARGs based on specific genomes
and their intervals of genetic inheritance. We contrast this ``gARG''
encoding with the classical approach derived from the coalescent
with recombination stochastic process of Griffiths
and colleagues~\citep{griffiths1991two,ethier1990two,
griffiths1996ancestral,griffiths1997ancestral}. As this is based
on recording evolutionary events, we refer to it as the event ARG, or
``eARG'' encoding.
We show that the gARG encoding is both flexible and efficient,
and can represent varying levels of precision about historical
events.
% Wilder: important to get this point into the intro
We argue that while adopting the general concept of an ARG data structure
is useful, it is important that we then fully decouple this data structure
from the original stochastic process.
[MORE]

@@

\section*{Genome ARGs}
% There is no fundamental reason, however, that the data structure we use
% to describe patterns of genetic inheritance must be in terms
% of specific causal events. As demonstrated in subsequent sections,
% there are significant advantages to using
% a slightly different formulation, which
% we call a ``genome ARG'' (gARG).
In this section we provide a  concrete definition of an ARG in terms
of particular genomes and their patterns of inheritance, which
we refer to as a ``genome ARG``, or gARG.
First, let us define a genome as the complete set of genetic material that a child
inherits from one parent (i.e., haploids have one genome, diploids two, etc.).
Then, a gARG is a graph in which nodes represent
monoploid genomes  and edges represent
genetic inheritance between an ancestor and a descendant,
where each edge is annotated with the coordinates
of the genomic intervals over which genetic inheritance occurs.
We can therefore view a gARG for a given set of individuals
as a subset of their pedigree, but, crucially,
where each individual is represented by two nodes (for diploids).
% To recover the local tree for a genomic
% position $x$, we traverse the graph rootwards from the leaves
% (as in an eARG) and at a particular node, we follow the outbound
% edge associated with the genomic interval containing $x$.
% Recent discussions have illustrated the
% relationship between ARGs and
% pedigrees~\citep{mathieson2020ancestry,brandt2021evaluation},
% and the deep connections between the structures are
% well understood
% \citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
% speed2015naturereviewsgenetics}.
This is illustrated in Fig.~\ref{fig-arg-in-pedigree},
which
shows how the genomes embedded in an example (highly inbred)
pedigree correspond to the nodes in a gARG.
In this diploid example
a genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination.

@@

There are some interesting consequences to these definitions, exemplified
in Figure~\ref{fig-arg-in-pedigree}. Firstly,
there is
no limitation on mating system, ploidy, or age structure. By making the
basic unit of analysis an individual's \emph{genomes} (i.e., two per individuals
for diploids), any pattern of inheritance among mixed ploidy can be
accommodated.
Secondly, because we only record the \emph{intervals} of inheritance
between genomes (and not the causal \emph{processes}),
arbitrary patterns of genetic inheritance such as
gene conversions,
multiple recombinations
across different chromosomes~\citep{fearnhead2003ancestral,koskela2019robust},
and so on can be directly expressed.
% Wilder: give forward ref to section on efficiency here, mention tree sequence

Thirdly, ``local'' trees that we generate for a particular position on the
genome are not explicitly concerned with ``coalescence'' (nodes at which
one or more samples find a common ancestor); for example, node $e$
is unary in the left-most tree of Figure~\ref{fig-arg-in-pedigree},
denoting the fact that $e$ the ancestor of $a$ over the interval $(0, 2]$,
but no coalescence occurs in this interval (see section XXX for more
discussion). Fourth, recombination is modelled in terms of
its outcomes in terms of inheritance intervals among parental genomes;
thus, we will have a child node (the recombinant; e.g. node $a$
Figure~\ref{fig-arg-in-pedigree}) and two (or more; see section XXX)
paternal genomes which this child has inherited segments from.
% TODO rephrase this, it'll confuse people. Just need to reiterate the
% points about events, and maybe lead into the next section
Finally, gARGs are not defined in terms of the
coalescent with recombination (or any other) stochastic process.
Thus, the gARG encoding is focused on faithfully representing the
\emph{outcome} of the evolutionary processes of genetic inheritance and
recombination, and is agnostic to the details of those processes.

@@

The gARG representation, on the other hand,
can be either retrospective or \emph{prospective}.
The edges in a gARG represent a path of genetic inheritance from
ancestor to descendant through some
% Wilder: finds the cell-to-cell confusing. Maybe drop?
number of generations (ultimately from cell to cell),
and are annotated
with the genome coordinates over which inheritance occurs.
In a prospective gARG, the inheritance intervals $I$
on an edge joining ancestor genome $v$ and descendant $u$
defines \emph{all} regions of genome that $u$ inherited from $v$.
Thus, these  inheritance intervals are
concerned only with the genetic relationship
between the nodes in question, and are not defined with respect
to any given set of samples. We are simply recording the
direct genetic inheritance between two genomes
(in the simplest case, between a parent and a child).

@@

The remaining nodes are MRCAs of some subset of the samples
at \emph{some} positions along the genome. We still have
some unary nodes in the local trees, but these nodes will
correspond to a coalescence in at least one other
local tree. For example, node  \noderef{k} is unary in the fourth tree
of Fig.~\ref{fig-simplification}C, but is either binary
or ternary in all subsequent trees (recall this is a Wright-Fisher
simulation). The final level of simplification is to alter the edge annotations
such that, although no nodes are removed from the graph, all
unary nodes disappear from the local trees (Fig.~\ref{fig-simplification}D).
Note that although this last stage produces simpler local trees, by
removing information about the exact paths taken by lineages through
the graph, we lose potentially useful information about shared edges
between trees. The extent to which this information can be inferred
and utilised is an interesting open question.

@@

\section*{Discussion}
Recent breakthroughs have made ARG inference methods practical
for the first time, and there has been a surge of interest
in inference methods, their evaluation and application.
The prospect of ARGs being used routinely in population
and statistical genetics applications is tantalising,
but in reality there is  substantial work to be done to
enable this.
